---
title: "Untitled"
author: "Julien Diot"
date: "2024-11-25"
output:
  ioslides_presentation:
    # css: src/style.css # not working within the nix package
    logo: src/imgs/IBS_IBC-Logo_2023.png
    widescreen: yes
    incremental: false
    smaller: false
    transition: faster
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hold" # display chunk output in one block
)

library(flair)
library(dplyr)
```
```{r css, results='asis'}
htmltools::includeCSS("src/style.css")
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with R Output

```{r echo=TRUE, eval=FALSE}
print("toto")
print("toto")
print("toto")
print("toto")
```

## Slide with R Output

<pre><code>
print("toto")
<strong>print("toto")</strong>
print("toto")
print("toto")
</code></pre>

## how_to_pipe

```{r how_to_pipe, include = FALSE}
print("toto")
print("tata")
print("titi")
print("tutu")
```


```{r, echo = FALSE}
decorate("how_to_pipe") %>%
  flair_lines(c(2, 4))
```


## Slide with Plot

```{r pressure}
plot(pressure)
```

