---
title: "Interactive Data Visualization in R with Shiny, Example of *PlantBreedGame*, A serious game to Teach Genomic Selection"
author: "Julien DIOT, The University of Tokyo, Lab. of Biometry and Bioinformatics"
date: "2024-12-10, IBC 2024, Atlanta USA"
output:
  ioslides_presentation:
    # css: src/style.css # not working within the nix package
    logo: src/imgs/IBS_IBC-Logo_2023.png
    widescreen: yes
    incremental: false
    smaller: false
    transition: faster
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results = "hold" # display chunk output in one block
)

library(qrcode)
library(plotly)
library(shiny)
```

```{r css, results='asis'}
htmltools::includeCSS("src/style.css")
```

## Overview

The aim of this presentation is to encourage you to try `Shiny` by showing
you simple example that you can reproduce and a more complex example
(*PlantBreedGame*) to show you `Shiny`'s capabilities.

- What is `Shiny`
- Simple `Shiny` examples (with `R`)
- Shiny application's structure
- Example of *PlantBreedGame*
- R-Shiny's Pro and con
- References



## What is `Shiny`

- Packages for building *"Easy **web apps for data science** without the compromises"* from `R` and `python`
- Developed by **Posit** (formly R-Studio)
- **`R`-Package** [github.com/rstudio/shiny](https://github.com/rstudio/shiny)
  - [Initial commit](https://github.com/rstudio/shiny/tree/e28b3da1badfecb34235e74a43aac4e8da1641bc)
  on Jun 21, 2012 by Joe Chang [@jcheng5](https://github.com/jcheng5)
  - [`v1.0.0`](https://github.com/rstudio/shiny/releases/tag/v1.0.0) published on Jan 14, 2017
  - ~88 contributors, ~5.4K ⭐, CRAN [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/last-month/shiny?color=blue)](https://r-pkg.org/pkg/shiny)
- **`Python` package** [github.com/posit-dev/py-shiny](https://github.com/posit-dev/py-shiny)
  - [first commit](https://github.com/posit-dev/py-shiny/tree/5f6905833822301a59757247140dce3f4d6cb339)
  on Jul 27, 2021 by Winston Chang [@wch](https://github.com/wch)
  - [`v1.0.0`](https://github.com/posit-dev/py-shiny/releases/tag/v1.0.0) published on Jul 19, 2024
  - ~28 contributors, ~1.3K ⭐, PiPy [![PyPI - Downloads](https://img.shields.io/pypi/dm/shiny)](https://pypistats.org/packages/shiny)


## Shiny example

- Basic example app from R-Studio: `File -> New File -> Shiny Web App...`

```{r}
ui <- fluidPage(
    # Application title
    titlePanel("Old Faithful Geyser Data"),

    # Sidebar with a slider input for number of bins
    sidebarLayout(
        sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30)
        ),

        # Show a plot of the generated distribution
        mainPanel(
           plotOutput("distPlot")
        )
    )
)

server <- function(input, output) {

    output$distPlot <- renderPlot({
        # generate bins based on input$bins from ui.R
        x    <- faithful[, 2]
        bins <- seq(min(x), max(x), length.out = input$bins + 1)

        # draw the histogram with the specified number of bins
        hist(x, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })
}

shinyApp(ui = ui, server = server)
```

## Shiny example, code

- Shiny app are structured around 2 components **`ui` and `server`**:


<div style="display: flex;">
<div style="flex: 1;">
<div style="text-align: center;">
**`ui`**
</div>

- defines **the visual** aspect of the application:
  - The general layout
  - The content
  - The inputs
  - The output types and location
</div>

<div style="flex: 1;">
<div style="text-align: center;">
**`server`**
</div>

  - defines **the logic to generate the outpus** from the inputs
</div>
</div>

- `shiny` handles the communication between `ui` and `server`
- The usage of inputs/outputs **"ids"** ensures **the links** between `ui` and `server`


## Shiny example, code  {.smaller}

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot")
    )
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  })
}
```

</div>
</div>

Then we can run start the app with `shinyApp(ui = ui, server = server)`

## Shiny example, code  {.smaller}

We define the page structure: 1 page (`fluidPage`) with a title (`titlePanel`), and a "sidebar" and "main area" layout (`sidebarLayout`, `sidebarPanel`, `mainPanel`)

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
  ### <b>
ui <- fluidPage(
  ### </b>
  # Application title
  ### <b>
  titlePanel("Old Faithful Geyser Data"),
  ### </b>

  # Sidebar with a slider input for number of bins
  ### <b>
  sidebarLayout(
    sidebarPanel(
  ### </b>
      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
  ### <b>
    ),
  ### </b>
    # Show a plot of the generated distribution
  ### <b>
    mainPanel(
  ### </b>
      plotOutput("distPlot")
  ### <b>
    )
  )
  ### </b>
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  })
}
```

</div>

</div>

## Shiny example, code  {.smaller}

We define the inputs: here **a slider** with `sliderInput` with **id `bins`**

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins

  sidebarLayout(
  ### <b>
    sidebarPanel(
      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
  ### </b>
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot")
    )
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  })
}
```

```{r}
sliderInput("bins_2", "Number of bins:", min = 1, max = 50, value = 30)
```

</div>

</div>




## Shiny example, code  {.smaller}

We define the outputs: here **a basic R plot** with `plotOutput` with **id `distPlot`**

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
    # Show a plot of the generated distribution
  ### <b>
    mainPanel(
      plotOutput("distPlot")
    )
  ### </b>
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  })
}
```
</div>
</div>


## Shiny example, code  {.smaller}

We specify the logic for the outputs `distPlot`.

- `plotOutput` works in combination with `renderPlot`

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
    # Show a plot of the generated distribution
    mainPanel(
  ### <b>
      plotOutput("distPlot")
  ### </b>
    )
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  ### <b>
  output$distPlot <- renderPlot({
  ### </b>
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  ### <b>
  })
  ### </b>
}
```
</div>
</div>


## Shiny example, code  {.smaller}

We access the value of the input with `input$<id>` here `input$bins`

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(

  ### <b>
      sliderInput("bins",
  ### </b>
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot")
    )
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
  ### <b>
                length.out = input$bins + 1)
  ### </b>

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  })
}
```
</div>
</div>

## Shiny example, code  {.smaller}

Inside the "render functions" it is standanrd R code that should return the correct object

<div style="display: flex;">

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
ui <- fluidPage(
  # Application title
  titlePanel("Old Faithful Geyser Data"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(

      sliderInput("bins",
        "Number of bins:",
        min = 1,
        max = 50,
        value = 30)
    ),
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("distPlot")
    )
  )
)
```
</div>

<div style="flex: 1;">

```{r, echo=TRUE, eval=FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({
  ### <b>
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2]
    bins <- seq(min(x), max(x),
                length.out = input$bins + 1)

    # draw the histogram with the specified
    # number of bins
    hist(x, breaks = bins, col = 'darkgray',
         border = 'white',
         xlab = 'Waiting time to next eruption (in mins)',
         main = 'Histogram of waiting times')
  ### </b>
  })
}
```
</div>
</div>


## Another simple example with `plotly`

`Shiny` works very well with other packages oriented on interactivity like [`plotly`](https://plotly.com/r/), [`DT`](https://rstudio.github.io/DT/), [`leaflet`](https://rstudio.github.io/leaflet/)...


<div class="sideBar_left">

```{r echo = FALSE, results='markup'}
data <- iris
columns <- colnames(data)

selectInput("x_var", "X variable", choices = columns, selected = columns[1])
selectInput("y_var", "Y variable", choices = columns, selected = columns[2])
```

</div>


<div class="sideBar_right">

```{r echo = FALSE, out.width = '10%', results='markup'}
renderPlotly({

  plot_data <- data[, c(input$x_var, input$y_var, "Species")]
  colnames(plot_data) <- c("x_var", "y_var", "color")

  p <- plot_ly(
    type = 'scatter',
    mode = 'markers',
    height = 400,
    width = 800,
    data = plot_data,
    x = ~x_var,
    y = ~y_var,
    color = ~color,
    hoverinfo = "text",
    text = apply(data, 1, function(l) { # data must be in the same order as plot_data here !
      paste(names(l), ":", l, collapse = "\n")
    })
  ) %>%
    layout(
      autosize = F,
      xaxis = list(title = input$x_var),
      yaxis = list(title = input$y_var)
    )

  p
})
```

</div>


## *PlantBreedGame*

- https://github.com/timflutre/PlantBreedGame

- Serious game to teach the principles of selective breeding build with `R-Shiny`
- Players are plant breeder and must carry out a successful selection campaign
- They upload "request files" and the app simulate the requested data: phenotypes, haplotypes genotypes

- App structured have been developed during a Master 2 student project in 2 months (some upates have been made after)


<div class="notes">
  http://127.0.0.1:3839

  The **main structure of the app** have been done during the 2 month project.
  Not the simulation code that were already developed.
  Other updates have been made later but most of the app structur was developed in 2 months
</div>


## Shiny Pros & Cons


<div style="display: flex;">
<div style="flex: 1;">
<div style="text-align: center;">
**Pros**
</div>
- **No webfront skills** required (but `HTML`/`CSS`/`JS` knowledge is helpfull)
- **Pure `R`** (or `Python`), any R-Package can be integrated
- Rapid development
- No web-server required
</div>

<div style="flex: 1;">
<div style="text-align: center;">
**Cons**
</div>
- Performances:
  - limited by **`R`'s performances**
  - single treaded: **Calculations "block" the app**, for all other users
- **Basic design** (without `HTML`/`CSS`/`JS`)
</div>
</div>

Summary:

- Great **for "statisticians"**
- Great for **internal communication** (eg. Shiny powered slides like those ones)
- Good for **Prototypes** / **Proof of Concept**, or **"Inernal App"**


<div class="notes">
  Statisticans ie. "R-developers" / "not web-dev"  
  Cons can be mitigated by specific R-Packages
</div>


## Some References { .smaller }

- `Shiny`'s official website: https://shiny.posit.co/
- Source code (for `R`): https://github.com/rstudio/shiny
- Source code (for `python`): https://github.com/posit-dev/py-shiny

Some usefull tools:

- `Shiny`'s gallery (several app examples): https://shiny.posit.co/r/gallery/
- List of basic UI inputs: https://gallery.shinyapps.io/081-widgets-gallery/
- Some `Shiny` extentions (for `R`): https://github.com/nanxstats/awesome-shiny-extensions

PlantBreedGame:

- Source code: https://github.com/timflutre/PlantBreedGame
- *Crop Science*, Letter to the Editor: *PlantBreedGame: A Serious Game that Puts Students in the Breeder's Seat*, by Timothée Flutre, Julien Diot, Jacques David, https://doi.org/10.2135/cropsci2019.03.0183le



---

<div style="text-align: center;">
**Thank you very much !**

Source code of the presentation is available on GitHub:  
https://github.com/juliendiot42/IBC_2024_presentation


```{r}
qr <- qr_code("https://github.com/juliendiot42/IBC_2024_presentation")
plot(qr)
```
</div>


<div class="notes">
  Links of refences in the repo
</div>

